CA Affordabulity

### 2020-09-21 paper.Rmd
update

---
title: "CA Affordabulity"
author: "Kyle Onda"
date: "9/3/2020"
output:
  html_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
library(sf)
library(beeswarm)
library(knitr)
library(stargazer)
```

@@

@@

tracts.bounds <- st_intersection(b.x,t.x) 
tracts.bounds$Area_TractPWS <- st_area(tracts.bounds)
tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract

bg.bounds <- st_intersection(bg.x,t.x) 
bg.bounds$Area_BlockGroupPWS <- st_area(bg.bounds)
bg.bounds$Percent_BlockGroupinPWS <- bg.bounds$Area_BlockGroupPWS/bg.bounds$AreaBlockGroup_Total

### Find 20% and 50% Income quintile ranges by CWS (use block groups)
tract.income <- tracts %>% filter(variable %in% c('hh_inc_5k',
                 'hh_inc_12.5k',
                 'hh_inc_17.5k',
                 'hh_inc_22.5k',
                 'hh_inc_27.5k',
                 'hh_inc_32.5k',
                 'hh_inc_37.5k',
                 'hh_inc_42.5k',
                 'hh_inc_47.5k',
                 'hh_inc_55k',
                 'hh_inc_67.5k',
                 'hh_inc_87.5k',
                 'hh_inc_112.5k',
                 'hh_inc_137.5k',
                 'hh_inc_175k',
                 'hh_inc_gr200k')) %>%
  st_drop_geometry() %>% group_by(GEOID) %>%
  mutate(cum_inc_sum = cumsum(estimate),
         percentile_inc = 100*cum_inc_sum/sum(estimate))

@@

@@

Overall, while 38% of community water systems are privately owned, they are disproportionately small systems, and so they only serve about 18% of the population. 94% of the population served by private utilities are served by utlities with more than 3,300 people. 
```{r boundaries, echo=FALSE,results='asis',message=FALSE, warning=FALSE}
b <- st_read("Data/boundaries.geojson") %>% filter(D_PWS_FED_=="C") 
b$size_cat= NA

b$size_cat[which(b$D_POPULATI<=500)] = "1. Very Small (<500)" 
b$size_cat[which(b$D_POPULATI>500)] = "2. Small (501-3300)" 
b$size_cat[which(b$D_POPULATI>3301)] = "3. Medium (3301-10,000)" 
b$size_cat[which(b$D_POPULATI>10001)] = "4. Large (10,001- 100,000)" 
b$size_cat[which(b$D_POPULATI>100000)] = "5. Very Large (>100,000)"

b$Ownership <-"Public"
b$Ownership[which(b$OWNER_TYPE=="P")]<-"Private"

size_summary <- b %>% group_by(size_cat)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()

size_summary$perc_pop <- 100*size_summary$pop/sum(size_summary$pop)
size_summary <- arrange(size_summary,size_cat)
kable(size_summary, caption= "Size and population distribution of CA Water Utilities",digits=1)

private <- b%>%group_by(Ownership)%>%add_tally()%>%summarize(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() %>% ungroup()%>%mutate(perc_pop=100*pop/sum(pop))
kable(private, caption= "Population served by public and private utilities in CA",digits=1)

size_summary.p <- b %>% group_by(size_cat,Ownership)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry()

size_summary.p$perc_pop <- 100*size_summary.p$pop/sum(size_summary.p$pop)
size_summary.p <- arrange(size_summary.p,size_cat,Ownership)
kable(size_summary.p, caption= "Size and population distribution and ownership of CA Water Utilities",digits=1)

@@

```{r rates pop,echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
rates <- read_csv("Data/affordability.csv")
priv <- select(b,PWSID,Ownership,D_POPULATI)
rates <- left_join(rates,b,by=c("pwsid"="PWSID")) 
rates <- rates[which(!is.na(rates$Ownership)),]
ggplot(rates, aes(x=log10(D_POPULATI), fill=Ownership)) +
  geom_density(alpha=0.4)
# Add mean lines
# ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
# ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
# ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
# ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
# ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
# ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
# 
# ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
# rm(ear13,ear14,ear15,ear16,ear17,ear18)
# save(ear,file="Data/EAR/ear.rds")
load("Data/EAR/ear.rds")
ear$QCat<-substr(ear$QuestionName,1,2)

@@

rates$Private <- 0
rates$Private[which(rates$Ownership=="Private")]<-1
ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
  geom_boxplot()

@@

- Are richer places having higher prices?
- Are private utiltiies tending to be in richer or poorer places than public utilities?

>>>>

---
title: "CA Affordabulity"
author: "Kyle Onda"
date: "9/3/2020"
output:
  html_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidycensus)
library(sf)
library(beeswarm)
library(knitr)
library(stargazer)
library(readxl)
```

@@

bounds <- st_read("Data/boundaries.geojson") %>% 
  filter(D_PWS_FED_=="C") %>% 
  select(PWSID,D_POPULATI,SVC_CONNEC)

@@

bg.x <- bg %>% filter(variable=="median_income") %>%
  select(GEOID,NAME,geometry) %>%
  st_transform(3310) %>%
  st_buffer(dist=0) %>%
  mutate(AreaBlockGroup = st_area(.))

tracts.bounds <- st_intersection(b.x,t.x) 
tracts.bounds$Area_TractPWS <- st_area(tracts.bounds)
tracts.bounds$Percent_TractinPWS <- tracts.bounds$Area_TractPWS/tracts.bounds$AreaTract
save(tracts.bounds,file="Data/tracts_bounds.rds")

bg.bounds <- st_intersection(b.x,bg.x) 
bg.bounds$Area_BlockGroupPWS <- st_area(bg.bounds)
bg.bounds$Percent_BlockGroupinPWS <- bg.bounds$Area_BlockGroupPWS/bg.bounds$AreaBlockGroup
save(bg.bounds,file="Data/block_groups_bounds.rds")

### Find 20% and 50% Income quintile ranges by CWS (use block groups)
pws.income <- bg.bounds %>% st_drop_geometry() %>%
  left_join(bg,by="GEOID") %>%
  filter(variable %in% c('hh_inc_5k',
                 'hh_inc_12.5k',
                 'hh_inc_17.5k',
                 'hh_inc_22.5k',
                 'hh_inc_27.5k',
                 'hh_inc_32.5k',
                 'hh_inc_37.5k',
                 'hh_inc_42.5k',
                 'hh_inc_47.5k',
                 'hh_inc_55k',
                 'hh_inc_67.5k',
                 'hh_inc_87.5k',
                 'hh_inc_112.5k',
                 'hh_inc_137.5k',
                 'hh_inc_175k',
                 'hh_inc_gr200k')) %>%
  group_by(PWSID,variable) %>%
  summarise(estimate=sum(estimate*Percent_BlockGroupinPWS)) %>% 
  ungroup() %>%
  mutate(var2=case_when(variable=="hh_inc_5k" ~ 1,
                 variable=="hh_inc_12.5k" ~ 2,
                 variable=="hh_inc_17.5k" ~ 3,
                 variable=="hh_inc_22.5k" ~ 4,
                 variable=="hh_inc_27.5k" ~ 5,
                 variable=="hh_inc_32.5k" ~ 6,
                 variable=="hh_inc_37.5k" ~ 7,
                 variable=="hh_inc_42.5k" ~ 8,
                 variable=="hh_inc_47.5k" ~ 9,
                 variable=="hh_inc_55k" ~ 10,
                 variable=="hh_inc_67.5k" ~ 11,
                 variable=="hh_inc_87.5k" ~ 12,
                 variable=="hh_inc_112.5k" ~ 13,
                 variable=="hh_inc_137.5k" ~ 14,
                 variable=="hh_inc_175k" ~ 15,
                 variable=="hh_inc_gr200k" ~ 16,)) %>% 
  arrange(PWSID,var2) %>% 
  group_by(PWSID) %>%
  mutate(cum_inc_sum = cumsum(estimate),
         percentile_inc = 100*cum_inc_sum/sum(estimate)) %>%
  ungroup() %>%
  mutate(perc_dist20=abs(as.numeric(percentile_inc)-20),
         perc_dist50=abs(as.numeric(percentile_inc)-50)) %>%
  group_by(PWSID) %>%
  filter(perc_dist20==min(perc_dist20) | 
         perc_dist50==min(perc_dist50)) %>%
  mutate(percentile=case_when(var2==min(var2)~"20th Percentile",
                              var2==max(var2)~"50th Percentile"),
         income_perc=case_when(var2==1 ~ 5000,
                               var2==2 ~ 12500,
                               var2==3 ~ 17500,
                               var2==4 ~ 22500,
                               var2==5 ~ 27500,
                               var2==6 ~ 32500,
                               var2==7 ~ 37500,
                               var2==8 ~ 42500,
                               var2==9 ~ 47500,
                               var2==10 ~ 55000,
                               var2==11 ~ 67500,
                               var2==12 ~ 87500,
                               var2==13 ~ 112500,
                               var2==14 ~ 137500,
                               var2==15 ~ 175000,
                               var2==16 ~ 225000)
         ) %>%
  pivot_wider(id_cols=PWSID,names_from=percentile,values_from=income_perc)

pws.income2 <-unnest(pws.income, cols = c(`20th Percentile`, `50th Percentile`, `NA`)) %>%
  select(PWSID,`20th Percentile`,`50th Percentile`) %>% distinct(.keep_all=TRUE)

save(pws.income2,file="Data/pws_income2.rds")

load("Data/own.rds")

@@

pws.rent <-  bg.bounds %>% 
  st_drop_geometry() %>%
  left_join(tracts,by="GEOID") %>%
  filter(variable=)

@@

Overall, while 38% of community water systems are privately owned, they are disproportionately small systems, and so they only serve about 18% of the population. 94% of the population served by private utilities are served by utlities with more than 3,300 people. 
<!-- # ```{r boundaries, echo=FALSE,results='asis',message=FALSE, warning=FALSE} -->
<!-- # b <- st_read("Data/boundaries.geojson") %>% filter(D_PWS_FED_=="C")  -->
<!-- # b$size_cat= NA -->
<!-- #  -->
<!-- # b$size_cat[which(b$D_POPULATI<=500)] = "1. Very Small (<500)"  -->
<!-- # b$size_cat[which(b$D_POPULATI>500)] = "2. Small (501-3300)"  -->
<!-- # b$size_cat[which(b$D_POPULATI>3301)] = "3. Medium (3301-10,000)"  -->
<!-- # b$size_cat[which(b$D_POPULATI>10001)] = "4. Large (10,001- 100,000)"  -->
<!-- # b$size_cat[which(b$D_POPULATI>100000)] = "5. Very Large (>100,000)"  -->
<!-- #  -->
<!-- # b$Ownership <-"Public" -->
<!-- # b$Ownership[which(b$OWNER_TYPE=="P")]<-"Private" -->
<!-- #  -->
<!-- # size_summary <- b %>% group_by(size_cat)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() -->
<!-- #  -->
<!-- # size_summary$perc_pop <- 100*size_summary$pop/sum(size_summary$pop) -->
<!-- # size_summary <- arrange(size_summary,size_cat) -->
<!-- # kable(size_summary, caption= "Size and population distribution of CA Water Utilities",digits=1) -->
<!-- #  -->
<!-- # private <- b%>%group_by(Ownership)%>%add_tally()%>%summarize(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() %>% ungroup()%>%mutate(perc_pop=100*pop/sum(pop)) -->
<!-- # kable(private, caption= "Population served by public and private utilities in CA",digits=1) -->
<!-- #  -->
<!-- # size_summary.p <- b %>% group_by(size_cat,Ownership)%>%add_tally()%>%summarise(count=mean(n),pop=sum(D_POPULATI)/10^6)%>%st_drop_geometry() -->
<!-- #  -->
<!-- # size_summary.p$perc_pop <- 100*size_summary.p$pop/sum(size_summary.p$pop) -->
<!-- # size_summary.p <- arrange(size_summary.p,size_cat,Ownership) -->
<!-- # kable(size_summary.p, caption= "Size and population distribution and ownership of CA Water Utilities",digits=1) -->
<!-- #  -->
<!-- #  -->
<!-- # ``` -->

@@

```{r rates pop,echo=FALSE, include=TRUE, message=FALSE,warning=FALSE, results='asis'}
rates <- read_csv("Data/affordability.csv")
rates <- left_join(rates,pws.income2,by=c("pwsid"="PWSID"))
rates$AR20 <- rates$total_bill_by_hhsize/rates$`20th Percentile`
rates$MHI <- rates$total_bill_by_hhsize/rates$`50th Percentile`

priv <- select(b,PWSID,Ownership,D_POPULATI)
rates <- left_join(rates,b,by=c("pwsid"="PWSID")) 
rates <- rates[which(!is.na(rates$Ownership)),]
ggplot(rates, aes(x=log10(D_POPULATI), fill=Ownership)) +
  geom_density(alpha=0.4)
# Add mean lines
# ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
# ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
# ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
# ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
# ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
# ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
# 
# ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
# rm(ear13,ear14,ear15,ear16,ear17,ear18)
# save(ear,file="Data/EAR/ear.rds")
load("Data/EAR/ear.rds")
ear$QCat<-substr(ear$QuestionName,1,2)

@@

rates$Private <- 0
rates$Private[which(rates$Ownership=="Private")]<-1
ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
  geom_boxplot() 
ggplot(rates, aes(x=as.factor(hh_sizes), y=total_bill_by_hhsize, colour=Ownership)) +
  geom_boxplot()

@@

- Are richer places having higher prices?

```{r richer higher}
prices <- select(own,
                 PWSID,
                 Year,
                 `Water System Classification`,
                 `Water System Ownership`,
                 `WR Type`,
                 `WR 6 HCF DW Charges`,
                 `WR 12 HCF DW Charges`) %>% filter(Year==2017 & `WR Type`!="-" & !is.na(`WR Type`))
prices.income <- inv %>% left_join(prices,by="PWSID") %>% left_join(pws.income2,by="PWSID")
```

- Are private utiltiies tending to be in richer or poorer places than public utilities?

```{r private richer}
